---
title: "Laboratory 03"
output: html_notebook
---

Carmen Martín Turrero

------------------------------------------------------------------------

Date : 14/04/2022

------------------------------------------------------------------------

```{r}
library(ggplot2)
library(GoFKernel)
library(latex2exp)
library(ggpubr)
```

------------------------------------------------------------------------

## Exercise 1

The triangular distribution, in the interval (a, b), is given by the following:

```{=tex}
\begin{equation}
f(x)=
\begin{cases}
\frac{2(x-a)}{(b-a)(c-a)} & a\leqslant x < c\\
\frac{2(b-x)}{(b-a)(b-c)} & c\leqslant x \leqslant b\\
0 & \text{otherwise}
\end{cases}
\end{equation}
```
where $c ∈ [a, b]$.

1.a. Plot the function, given the interval (a, b).

```{r}

a<- 1
b <- 3
c <- 2.5

d.triangular <- function(x) {
  if (a<=x & x<c) {
    d <- 2*(x-a)/((b-a)*(c-a))
  } else if (c<= x & x <= b) {
    d<- 2*(b-x)/((b-a)*(b-c))
  } else {
    d <- 0
  } 
  return (d)
}

d.triangular <- Vectorize(d.triangular)

```

```{r}

X <- seq(0, 4, 4/100)
tri <- d.triangular(X)
plot_df <- data.frame(X, tri)

ggplot() +
  geom_line(data = plot_df, aes(x=X, y=tri), linetype =2 , size = 0.8) +
  labs(title = "Triangular distribution", subtitle = paste("a =", a, ", b = ", b, ", c =", c), x = "x", y = "pdf")+
  guides(color=guide_legend(NULL, order = 5))

```

1.b. Write an algorithm to generate random numbers from the triangular distribution.

```{r}

r.triangular <- function(N, a, b, c){
  x <- seq(a,b, (b-a)/2000)
  d.triangular <- function(x) {
    ifelse((x>a & x<c), (2*(x-a))/((b-a)*(c-a)),
           ifelse((x>=c & x<b), (2*(b-x))/((b-a)*(b-c)),0))
  }
  
  p.triangular <- function(x) {
    ifelse((x>a & x<b), integrate(d.triangular,a,x)$value,
           ifelse(x <= a, 0, 1))
  }
  
  sampl <- runif(N)
  rf <- Vectorize(inverse(p.triangular, a, b))
  return (rf(sampl))
}



```


1.c. Generate $10^4$ random numbers from the distribution, show them in an histogram and superimpose the analytical curve.

```{r}

samples <- r.triangular(10000, a,b,c)


```

```{r}
samples.df <- data.frame(samples)

ggplot() +
  geom_histogram(data = samples.df, aes(x=samples, y=..density..), color="darkblue", fill="lightblue", size = 0.5) +
  geom_line(data = plot_df, aes(x=X, y=tri, color = "PDF"), linetype =2 , size = 0.8) +
  labs(title = "Triangular distribution", subtitle = paste("a =", a, ", b = ", b, ", c =", c), x = "x", y = "pdf")+
  guides(color=guide_legend(NULL, order = 5))

```

------------------------------------------------------------------------

## Exercise 2 - Markov's inequality

Markov's inequality represents an upper bound to probability distributions:

```{=tex}
\begin{equation}
P(X\geqslant k)\leqslant \frac{E[X]}{k}, for k>0.
\end{equation}
```
Having defined a function $G(k) = 1 - F(k) \equiv P(X\geqslant k)$ plot G(k) and the Markov's upper bound for

2.a. The exponential, Exp(λ = 1), distribution function.

```{r}

# Function G(k)
G.exp  <- Vectorize(function(k){1- pexp(k, 1)})

#Expected value of exponential with lambda = 1
E.exp <- 1

# Function for upper bound
up.bound <- Vectorize(function(e, k){e/k})

# Choose different k values and plot
k <- seq(0.5, 11.5, 0.5)
exp.df <- data.frame(k = k, E = up.bound(E.exp,k), G = G.exp(k))

ggplot() +
  geom_line(data = exp.df, aes(x = k, y = E, color = "Upper bound"), linetype = 2) +
  geom_line(data = exp.df, aes(x = k, y = G, color = "G(k)")) +
  ggtitle ("Exponential distribution, exp(1)") +
  xlab("x") + ylab(NULL) +
  guides(color=guide_legend(NULL, order = 5)) + 
  scale_colour_brewer(palette = "Dark2")
```

2.b. The uniform, U(3, 5), distribution function

```{r}

# Function G(k)
G.unif  <- Vectorize(function(k){1- punif(k, 3, 5)})

#Expected value of exponential with lambda = 1
E.unif <- 4

# Choose different k values and plot
k <- seq(0.5, 11.5, 0.5)
unif.df <- data.frame(k = k, E = up.bound(E.unif,k), G = G.unif(k))

ggplot() +
  geom_line(data = unif.df, aes(x = k, y = E, color = "Upper bound"), linetype = 2) +
  geom_line(data = unif.df, aes(x = k, y = G, color = "G(k)")) +
  ggtitle ("Uniform distribution, U(3, 5)") +
  xlab("x") + ylab(NULL) +
  guides(color=guide_legend(NULL, order = 5)) + 
  scale_colour_brewer(palette = "Dark2")

```

2.c. The binomial, Bin(n = 1, p = 1/2), distribution function

```{r}

# Function G(k)
G.binom  <- Vectorize(function(k){1- pbinom(k, 1, 0.5)})

#Expected value of exponential with lambda = 1
E.binom <- 0.5

# Choose different k values and plot
k <- seq(0.1, 11.5, 0.1)
binom.df <- data.frame(k = k, E = up.bound(E.binom,k), G = G.binom(k))

ggplot() +
  geom_line(data = binom.df, aes(x = k, y = E, color = "Upper bound"), linetype = 2) +
  geom_line(data = binom.df, aes(x = k, y = G, color = "G(k)")) +
  ggtitle ("Binomial distribution, bin(n = 1, p = 0.5)") +
  xlab("x") + ylab(NULL) +
  guides(color=guide_legend(NULL, order = 5)) + 
  scale_colour_brewer(palette = "Dark2") +
  coord_cartesian(ylim=c(0, 5))


```

2.d. a Poisson, Pois(λ = 1/2), distribution function

```{r}

# Function G(k)
G.pois  <- Vectorize(function(k){1- ppois(k, 0.5)})

#Expected value of exponential with lambda = 1
E.pois <- 0.5

# Choose different k values and plot
k <- seq(0.5, 11.5, 0.5)
pois.df <- data.frame(k = k, E = up.bound(E.pois,k), G = G.pois(k))

ggplot() +
  geom_line(data = pois.df, aes(x = k, y = E, color = "Upper bound"), linetype = 2) +
  geom_line(data = pois.df, aes(x = k, y = G, color = "G(k)")) +
  ggtitle ("Poisson Distribution") +
  xlab("x") + ylab(NULL) +
  guides(color=guide_legend(NULL, order = 5)) + 
  scale_colour_brewer(palette = "Dark2")

```

------------------------------------------------------------------------

## Exercise 3 - Chebyshev's inequality

Chebyshev's inequality tells us that

```{=tex}
\begin{equation}
P(|X-\mu|\geqslant k\sigma)\leqslant \frac{1}{k^2}
\end{equation}
```
which can also be written as

```{=tex}
\begin{equation}
P(|X-\mu|< k\sigma)\geqslant 1-\frac{1}{k^2}
\end{equation}
```
use R to show, with a plot, that Chebyshev's inequality is an upper bound to the following distributions:

3.a. A normal distribution, $N(µ = 3, σ = 5).$

```{r}

# Define Chebyshevs lower bound
low.bound <- Vectorize(function(k){1 - 1/(k*k)})

# Define mean and variance for normal distribution
mu <- 3         
sigma <- 5

P.norm <- Vectorize(function(k){pnorm(k*sigma + mu, mu, sigma) - pnorm(mu-k*sigma, mu, sigma)})

# Choose different k values and plot
k <- seq(0.5, 11.5, 0.5)
norm.df <- data.frame(k = k, low = low.bound(k), P = P.norm(k))

ggplot() +
  geom_line(data = norm.df, aes(x = k, y = low, color = "Lower bound"), linetype = 2) +
  geom_line(data = norm.df, aes(x = k, y = P, color = "P(k)")) +
  ggtitle ("Normal Distribution, N(3, 5)") +
  xlab("x") + ylab(NULL) +
  guides(color=guide_legend(NULL, order = 5)) + 
  scale_colour_brewer(palette = "Set1")

```

3.b. An exponential distribution, $Exp(λ = 1).$

```{r}

# Define the probability
P.exp <- Vectorize(function(k){pexp(1 + k, 1) - pnorm(1-k, 1)})

# Choose different k values and plot
k <- seq(0.5, 11.5, 0.5)
exp.df <- data.frame(k = k, low = low.bound(k), P = P.exp(k))

ggplot() +
  geom_line(data = exp.df, aes(x = k, y = low, color = "Lower bound"), linetype = 2) +
  geom_line(data = exp.df, aes(x = k, y = P, color = "P(k)")) +
  ggtitle ("Exponential Distribution") +
  xlab("x") + ylab(NULL) +
  guides(color=guide_legend(NULL, order = 5)) + 
  scale_colour_brewer(palette = "Set1")

```

3.c. A uniform distribution $U(1 − √2, 1 + √2).$

```{r}

# Define the probability
E.unif <- 1        
s.unif <- sqrt(2/3)

P.unif <- Vectorize(function(k){punif(E.unif + k*s.unif, 1-sqrt(2), 1+sqrt(2)) - 
    punif(E.unif - k*s.unif, 1-sqrt(2), 1+sqrt(2))})

# Choose different k values and plot
k <- seq(0.5, 11.5, 0.5)
unif.df <- data.frame(k = k, low = low.bound(k), P = P.unif(k))

ggplot() +
  geom_line(data = unif.df, aes(x = k, y = low, color = "Lower bound"), linetype = 2) +
  geom_line(data = unif.df, aes(x = k, y = P, color = "P(k)")) +
  ggtitle ("Exponential Distribution") +
  xlab("x") + ylab(NULL) +
  guides(color=guide_legend(NULL, order = 5)) + 
  scale_colour_brewer(palette = "Set1")

```

3.d. A Poisson, $Pois(λ = 1/3)$, distribution function.

```{r}

# Define the probability
l = 1/3
P.pois <- Vectorize(function(k){ppois(l + k*sqrt(l), l) - ppois(l - k*sqrt(l), l)})

# Choose different k values and plot
k <- seq(0.5, 11.5, 0.5)
pois.df <- data.frame(k = k, low = low.bound(k), P = P.pois(k))

ggplot() +
  geom_line(data = pois.df, aes(x = k, y = low, color = "Lower bound"), linetype = 2) +
  geom_line(data = pois.df, aes(x = k, y = P, color = "P(k)")) +
  ggtitle ("Exponential Distribution") +
  xlab("x") + ylab(NULL) +
  guides(color=guide_legend(NULL, order = 5)) + 
  scale_colour_brewer(palette = "Set1")


```

------------------------------------------------------------------------

## Exercise 4 - Six Boxes Toy Model: inference

Write a program that:

1.  Allows the user to insert the color of a randomly extracted box
2.  prints on the standard output the probability of selecting each box
3.  Plots the probability for each box as a function of the extraction step

```{r}
#Remember: H_0 = all black, H_5 = all white

# Before any extraction the probability of having selected each box is
prob_labels <- c('p_H0', 'p_H1', 'p_H2', 'p_H3', 'p_H4', 'p_H5')
probs_0 <- c(1/6, 1/6, 1/6, 1/6, 1/6, 1/6)

# The probability of extracting a black stone given the box is
pblack_bybox <- c(1, 4/5, 3/5, 2/5, 1/5, 0)

# The probability of extracting a black or white stone at the beginning is 1/2,
# since there is the same number of black and white stones.
p_E <- 0.5

# Having this data, we define a function that estimates the probability of
# having selected each box

estimation <- function(extraction, previous_estimate){
    estimate <- previous_estimate
    # black extraction
    if (extraction == 0) {
        #redefine the probability of extracting a black stone
        p_E <- sum(pblack_bybox * previous_estimate)
        estimate <- (previous_estimate*pblack_bybox)/p_E
    }
    #white extraction
    if (extraction == 1) {
        #redefine the probability of extracting a black stone
        p_E <- sum((1-pblack_bybox) * previous_estimate)
        estimate <- (previous_estimate*(1-pblack_bybox))/p_E
    }
    return (estimate)
}
```

```{r}

# function that plots probabilities for each extraction

probs.plot <- function(trials, data0, data1, data2, data3, data4, data5){
  probs.df <- data.frame(trials = trials, H0 = data0, H1 = data1, H2 = data2, H3 = data3, H4 = data4, H5 = data5)

  H0 <- ggplot() +
    geom_point(data = probs.df, aes(x = trials, y = H0), col = 'brown1')+
    ggtitle ("Box H0") + coord_cartesian(ylim = c(0,1)) +
    xlab("Trial") + ylab("Probability") +
    scale_colour_brewer(palette = "Set2")

  H1 <- ggplot() +
    geom_point(data = probs.df, aes(x = trials, y = H1), col = 'brown1')+
    ggtitle ("Box H1") + coord_cartesian(ylim = c(0,1)) +
    xlab("Trial") + ylab("Probability") +
    scale_colour_brewer(palette = "Set2")
  
  H2 <- ggplot() +
    geom_point(data = probs.df, aes(x = trials, y = H2), col = 'brown1')+
    ggtitle ("Box H2") + coord_cartesian(ylim = c(0,1)) +
    xlab("Trial") + ylab("Probability") +
    scale_colour_brewer(palette = "Set2")
  
  H3 <- ggplot() +
    geom_point(data = probs.df, aes(x = trials, y = H3), col = 'brown1')+
    ggtitle ("Box H3") + coord_cartesian(ylim = c(0,1)) +
    xlab("Trial") + ylab("Probability") +
    scale_colour_brewer(palette = "Set2")
  
  H4 <- ggplot() +
    geom_point(data = probs.df, aes(x = trials, y = H4), col = 'brown1')+
    ggtitle ("Box H4") + coord_cartesian(ylim = c(0,1)) +
    xlab("Trial") + ylab("Probability") +
    scale_colour_brewer(palette = "Set2")
  
  H5 <- ggplot() +
    geom_point(data = probs.df, aes(x = trials, y = H5), col = 'brown1')+
    ggtitle ("Box H5") + coord_cartesian(ylim = c(0,1)) +
    xlab("Trial") + ylab("Probability") +
    scale_colour_brewer(palette = "Set2")
  
  ggarrange(H0, H1, H2, H3, H4, H5,
            ncol = 3, nrow = 2)
}

```

```{r}
#Set the initial conditions
extraction  <- 0 
counter <- 0
previous <- probs_0
data0 <- probs_0[1]
data1 <- probs_0[2]
data2 <- probs_0[3]
data3 <- probs_0[4]
data4 <- probs_0[5]
data5 <- probs_0[6]
trials <- 0

while(extraction != 2){
  # Ask the user to insert the color of the extracted stone  
  extraction <- as.integer(readline(prompt="Which stone did you extract? (black = 0, white = 1, exit = 2) \n"))
  
  # Estimate the probabilities of having selected each box
  probs <- estimation(extraction, previous)
  counter <- counter + 1
  
  # Save the data for plotting
  data0 <- append(data0, probs[1])
  data1 <- append(data1, probs[2])
  data2 <- append(data2, probs[3])
  data3 <- append(data3, probs[4])
  data4 <- append(data4, probs[5])
  data5 <- append(data5, probs[6])
  trials <- append(trials, counter)
  
  # Print the actual probabilities
  cat('Trial ', counter,': P(H0) = ', round(probs[1], 8), ', P(H1) = ', round(probs[2], 8), ', P(H2) = ', round(probs[3],8), ', P(H3) = ', round(probs[4], 8), ', P(H4) = ', round(probs[5], 8), ', P(H5) = ', round(probs[6],8))
  
  # If we want to plot at each iteration use:
  # probs.plot(trials, data0, data1, data2, data3, data4, data5)
  
  # Assign the new values to use in the next iterations as previous ones
  previous <- probs
}    

if ( extraction == 2) {
  print('End of experiment')
}

```

```{r}

# Print final result (last trial)
probs.plot(trials, data0, data1, data2, data3, data4, data5)

```


------------------------------------------------------------------------

## Exercise 5 - Six Boxes Toy Model: simulation

Consider again the six boxes toy model of the previous exercise and write a simulation program that: 1) selects a random box 2) makes random sampling from the box 3) prints on the standard output the probability of selecting each box 4) plots the probability for each box as a function of the number of trial

```{r}

# The initial conditions are the same as in the previous exercise
# We will also use the functions defined in that exercise

box_labels <- c('H0', 'H1', 'H2', 'H3', 'H4', 'H5')
probs_0 <- c(1/6, 1/6, 1/6, 1/6, 1/6, 1/6)
pblack_bybox <- c(1, 4/5, 3/5, 2/5, 1/5, 0)
p_E_0 <- 0.5

# We select a box at random:
box.index <- sample(1:6, 1)

message("Selected box is: ", box_labels[box.index])

```

```{r}
#Make a random sampling from the box:

samples <- rbinom(100, 1, 1-pblack_bybox[box.index])
```


```{r}
# Study the probability of having selected each box based on the sampling
counter <- 0
previous <- probs_0
data0 <- probs_0[1]
data1 <- probs_0[2]
data2 <- probs_0[3]
data3 <- probs_0[4]
data4 <- probs_0[5]
data5 <- probs_0[6]
trials <- 0

for (extraction in samples){
  
  # Estimate the probabilities of having selected each box
  probs <- estimation(extraction, previous)
  counter <- counter + 1
  
  # Save the data for plotting
  data0 <- append(data0, probs[1])
  data1 <- append(data1, probs[2])
  data2 <- append(data2, probs[3])
  data3 <- append(data3, probs[4])
  data4 <- append(data4, probs[5])
  data5 <- append(data5, probs[6])
  trials <- append(trials, counter)
  
  # Assign the new values to use in the next iterations as previous ones
  previous <- probs
}
```


```{r}
# Print the final values and plot the estimation over time
cat('P(H0) = ', round(probs[1], 8), ', P(H1) = ', round(probs[2], 8), ', P(H2) = ', round(probs[3],8), ', P(H3) = ', round(probs[4], 8), ', P(H4) = ', round(probs[5], 8), ', P(H5) = ', round(probs[6],8))

probs.plot(trials, data0, data1, data2, data3, data4, data5)

```
